---
title: "Homework Assignment 2: Exploring patterns of environmental justice"
author: "Caitlin Nordheim-Maestas"
date: "`r Sys.Date()`"
format: html
---

Instructions here: <https://eds-223-geospatial.github.io/assignments/HW2.html>

HINT: LOOK UP CRS IN METADATA!

# Setup

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(sf)
library(tmap)
library(here)
library(viridisLite) # colors
library(janitor)
library(kableExtra) # pretty table

# load in data 
#ejscreen
ej <- st_read(here::here("data", "ejscreen", "EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb"))
# check valid
unique(st_is_valid(ej)) # yay

# holc
holc <- st_read(here::here("data", "mapping-inequality", "mapping-inequality-los-angeles.json")) %>% 
  st_make_valid()

# check and make valid
unique(st_is_valid(holc)) # yay!


# bird
bird <- st_read(here::here("data", "gbif-birds-LA", "gbif-birds-LA.shp"))
# check and make valid
unique(st_is_valid(bird)) # yay!
```


# Part 1

# Part 1.1

Create a map of historical redlining neighborhoods

```{r}
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 6

m1.1 <- tm_shape(holc) + # import holc geo data
  tm_basemap("Esri.WorldGrayCanvas") + # dull background that still has geo info
  tm_graticules() + # to get our bearings
  # fill by HOLC "grade" assigned
  tm_fill(fill = "grade", # each hold grade a different color
          fill.scale = tm_scale(values =  # assigning colors manually to match the historical figures
                                  c("green", "blue", "yellow", "pink", "black")),
          # this matches historical redlining maps in the metadata
          fill.legend = tm_legend(title = "HOLC Grade")) +
   tm_borders(col = "black", lwd = 0.5) + # makes it look more refined with borders
  tm_title("HOLC Grade Designations in Los Angeles") + # add title
# add map elements
  tm_compass(show.labels = 1, position = c("right","top")) + # compas
  tm_scalebar(position = c("right", "bottom")) + # scalebar
  tm_credits("Data Source: Digital Scholarship Lab", position = c("right", "bottom")) # data credits

m1.1
```

## Part 1.2

Create a table summarizing:

- the percentage of census block groups that fall within each HOLC grade

- Also include the percent of census black groups that don’t fall within a HOLC grade

- Hint: The HOLC data contains the grades and the EJScreen data contains the census blocks, so you will need to combine the data spatially before doing summary statistics. Once you combine and no longer need the geometries, you can use st_drop_geometry().

```{r}
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 6

# filter ej data to just LA county
ej.la <- ej %>%
  filter(CNTY_NAME == "Los Angeles County")

# crs check!!
# st_crs(ej.la) == st_crs(holc) # false
ej.la <- st_transform(ej.la, crs= st_crs(holc)) # set to same as holc
# st_crs(ej.la) == st_crs(holc) # true, yay!

# JOIN 
# join x,y adds y columns to x
ej_holc <- st_join(ej.la, holc) %>%
  clean_names() %>% # bc the caps are driving me crazy
  st_drop_geometry() # no longer needed

### summary stats
# the percentage of "census block groups" that fall within each HOLC grade
# percent of census block groups that don’t fall within a HOLC grade
# total rows #nrow(ej_holc) # 8988
ej_holc %>% 
  group_by(grade) %>% 
  # number in this group/ total rows *100 %
  #rounded to 2 decimal places
 summarise(
    Percent = round((n() / 8988) * 100, 2)) %>% 
  mutate(Percent = paste0(Percent, "%")) %>% 
  st_drop_geometry() %>% 
  # pretty table
  kbl(booktabs = T, caption = "Census block groups and percent") %>%
  kable_styling(latex_options = "striped")
```







